---
title: "Dataset Description and PCA Analysis"
author: "Your Name"
output:
  pdf_document:
    latex_engine: xelatex
date: "`r Sys.Date()`"
---

# 1. Dataset Description

## 1.1 Overview

The dataset contains multivariate data, including quantitative, binary, and categorical variables. It has 12,330 rows and 18 columns, representing user online shopping behavior. \## 1.2 Feature Breakdown

Below is a table representation with all the key features:

| **Feature Name**        | **Type**     | **Description**                                                   |
|-------------------|-----------------|------------------------------------|
| Administrative          | Quantitative | Number of administrative pages visited                            |
| Administrative_Duration | Quantitative | Time spent on administrative pages in seconds                     |
| Informational           | Quantitative | Number of informational pages visited                             |
| Informational_Duration  | Quantitative | Time spent on informational pages in seconds                      |
| ProductRelated          | Quantitative | Number of product-related pages visited                           |
| ProductRelated_Duration | Quantitative | Time spent on product-related pages in seconds                    |
| BounceRates             | Quantitative | Percentage of visitors leaving after one page                     |
| ExitRates               | Quantitative | Percentage of sessions exiting from each page                     |
| PageValues              | Quantitative | Average value attributed to a page                                |
| SpecialDay              | Quantitative | Metric indicating proximity to significant holidays               |
| Weekend                 | Binary       | Indicates if the session occurred on a weekend (1 = Yes, 0 = No)  |
| Revenue                 | Binary       | Indicates if the session resulted in a purchase (1 = Yes, 0 = No) |
| VisitorType             | Categorical  | Visitor category (e.g., Returning, New, Other)                    |
| Month                   | Categorical  | Month of the visit (e.g., Jan, Feb)                               |
| OperatingSystems        | Categorical  | Visitor\u2019s operating system                                   |
| Browser                 | Categorical  | Browser used by the visitor                                       |
| Region                  | Categorical  | Visitor\u2019s geographical region                                |
| TrafficType             | Categorical  | Type of traffic source leading to the visit                       |

## 1.3 Initial Observations

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# Load necessary libraries
library(ggplot2)   # For data visualization
library(readr)     # For reading datasets
library(lattice)   # For lattice-based visualizations
library(reshape2)  # For reshaping data
library(dplyr)     # For data manipulation
library(pander)
```

```{r}
data <- readRDS('/Users/despoinaiapona/Downloads/processed_dataset.rds')

# Checking for missing values
sapply(data, function(x) sum(is.na(x)))
```
\vspace{0.3cm}


```{r}
# Split numeric and categorical variables
numeric_vars <- data %>% select(where(is.numeric))
categorical_vars <- data %>% select(where(is.factor))
```

```{r}
# Summarize numeric variables
numeric_summary <- summary(numeric_vars)

# Display the summary
pander(numeric_summary, caption = "Summary of Numeric Variables")

```

```{r}
# Summarize categorical variables
categorical_summary <- summary(categorical_vars)

# Display the summary
pander(categorical_summary, caption = "Summary of Categorical Variables")
```
\vspace{1.5cm}

## 1.4 Visualizations

### 1.4.1  Barplots for Categorical Variables

```{r}
# Set the layout for 3 rows and 3 columns (9 plots per page)
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))  # Adjust margins

# Create barplots
for (var_name in names(categorical_vars)) {
  barplot(table(categorical_vars[[var_name]]), 
          main = paste("Distribution of", var_name), 
          xlab = var_name, 
          ylab = "Count", 
          col = "lightblue")
}
```


#### 1.4.1.1 General Observations

-   Most sessions happen on regular days, with little activity on holidays.

-   There are seasonal patterns, with high points in May and November.

-   Most users go with Operating System 2 and Browser 2.

-   Traffic comes mostly from Region 1 and Traffic Type 2.

-   Returning visitors account for a significant part of site sessions, indicating that users enjoy the content and experience.

-   The majority of sessions are on weekdays, and only a small part lead to sales.

\vspace{0.3cm}


```{r}
# Set layout for 3 rows and 3 columns (9 plots per page)
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))  # Adjust margins

# Create histograms for each numeric variable
for (var_name in names(numeric_vars)) {
  hist(numeric_vars[[var_name]], 
       main = paste("Histogram of", var_name), 
       xlab = var_name, 
       col = "lightgreen", 
       breaks = 20)
}
```



```{r}
for (i in 1:ncol(numeric_vars)) {
  var_name <- colnames(numeric_vars)[i]
  
  par(mfrow = c(1, 2), mar = c(5, 4, 4, 2) + 0.1, oma = c(0, 0, 2, 0))
  
  # Boxplot
  boxplot(numeric_vars[[i]], horizontal = TRUE, 
          main = paste("Boxplot of", var_name), 
          xlab = "Values")
  
  # Histogram
  hist(numeric_vars[[i]], breaks = 20, 
       main = paste("Histogram of", var_name), 
       xlab = var_name, 
       ylab = "Frequency")
  
  # Add mean and median to the title
  mtext(paste("Variable:", var_name, "| Mean:", round(mean(numeric_vars[[i]]), 2), 
              "| Median:", round(median(numeric_vars[[i]]), 2)), 
        outer = TRUE, cex = 1, line = -1)
}
```

```{r}
# Compute the correlation matrix
correlation_matrix <- cor(numeric_vars)

correlation_melted <- melt(correlation_matrix)

# Heatmap of the correlation matrix with values
ggplot(correlation_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +  # Create the heatmap tiles
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +  # Add the correlation values inside the tiles
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +  # Color scale
  labs(title = "Correlation Matrix Heatmap", x = "Variables", y = "Variables") +  # Title and labels
  theme_minimal() +  # Clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-ax

```

```{r}
# PCA with R built in function
pca <- prcomp(numeric_vars, center = TRUE, scale. = TRUE)
pca$rotation
```

```{r}
# PCA step by step

eig_result <- eigen(correlation_matrix)

# Get the eigenvalues and eigenvectors
Lambda <- eig_result$values  # Eigenvalues
T <- eig_result$vectors      # Eigenvectors

# Sort eigenvalues and eigenvectors in descending order
sorted_indices <- order(Lambda, decreasing = TRUE)  # Indices for sorting

# Apply the sorting
Lambda_sorted <- Lambda[sorted_indices]  # Sorted eigenvalues
T_sorted <- T[, sorted_indices]          # Sorted eigenvectors

# Show the sorted eigenvalues
Lambda_sorted

# Show the eigenvectors
T_sorted



```
